Nathan Greenslit Data Analysis Project

### 2023-04-06 code/analysis_code/env/env_par.qmd
Addedto env data viz

```{r}
daily<- d1 %>%
  filter(date %in% c(as.Date("2022-07-07"):as.Date("2022-07-19"))) %>%
  select(!...1)
```

@@

#### Plotting Daily Temperature

```{r}
temp<- daily %>% ggplot() + geom_line(
  aes(
    x = date,
    y = temp,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Temperature (°C)",
    title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 15),
    legend.position = c(0.1,0.2),
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))

temp

```

#### Plotting Daily Salinity

```{r}
sal<- daily %>% ggplot() + geom_line(
  aes(
    x = date,
    y = sal,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Salinity",
    title = "Salinity across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_text(size = 15),
    axis.title.y = element_text(size = 15),
    axis.text.x = element_text(size = 10),
    axis.text.y = element_text(size = 10),
    legend.position = "none")+
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))

sal
```

#### Combine Temperature and Salinity Plots

```{r, fig.width= 4, fig.height=4}
fig<- ggarrange(temp,sal + font("x.text", size = 10), ncol = 1, nrow = 2,align = "v")

@@

#### Save as PNG

>>>>

```{r}
daily<- d1 %>%
  filter(date %in% c(as.Date("2022-07-07"):as.Date("2022-07-19"))) %>%
  filter(!date %in% as.Date('2022-07-10')) %>%
  select(!...1)
```

@@

#### Temperature

```{r}
temp<- daily %>% ggplot() + geom_line(
  aes(
    x = date,
    y = temp,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Temperature (°C)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Salinity

```{r}
sal<- daily %>% ggplot() + geom_line(
  aes(
    x = date,
    y = sal,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Salinity",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Dissolved Oxygen

```{r}
do<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = do_mgl,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "DO (mg/L)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Nitrite:Nitrate

```{r}
nn<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = nn,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Nitrate-Nitrite (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Orthophosphate

```{r}
ortho<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = orthop,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Orthophosphate (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### DON

```{r}
don<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = don,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "DON (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### DOC

```{r}
doc<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = doc,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "DOC (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.ticks.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### pH

```{r}
pH<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = pH,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "pH",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_text(size = 10),
    axis.text.x = element_text(size = 10, angle = 45, hjust = 1),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Ammonium

```{r}
amm<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = amm,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Ammonium (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Total Nitrogen

```{r}
tn<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = tn,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Total Nitrogen (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Silicate

```{r}
sil<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = sil,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Silicate (µM)",
  #  title = "Temperature across Daily Time Series by Site",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.position = "none",
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid')) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Copies/mL

```{r}
cop<-
  daily %>%
  ggplot() + geom_line(
  aes(
    x = date,
    y = copies_mL,
    color = site),
  size = 1)+
  theme_bw()+
  labs(
    x = "Date",
    y = "Vibrio (copies per mL)",
    color = "Site")+
  theme(
    plot.title= element_text(hjust = 0.5, size = 20),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 7),
    axis.text.y = element_text(size = 7),
    legend.key.size = unit(0.3,'cm'),
    legend.position = c(0.2,0.7),
    legend.background = element_rect(colour = 'black', fill = 'white', linetype='solid'),
    legend.text = element_text(size = 7)) + #Moves legend onto plot 
    scale_color_manual(values = c("Blind Oso" = "#809C52",
                                "Canals" = "#ba4716",
                                "Gulf" = "4c848c"))
```

#### Combine Environment Plots

```{r, fig.width= 10, fig.height= 5}
fig<- ggarrange(cop,temp,sal,sil,do,amm,nn,tn,ortho,doc, pH,don + font("x.text", size = 10), ncol = 4, nrow = 3,align = "v")

@@

# Save as PNG

### 2023-04-06 products/manuscript/Manuscript.qmd
Addedto env data viz

3.  Ultimately, I feel as though I am missing some "oomph"when it comes to my statistical analysis. The linear models seem tadequate, but I know that there are multiple variables at play at these sites (and we have the nutrient data for this). I am open to any suggestions that may diversify and expand my analyses!

@@

## General Background Information

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeidaModelSaharanDust1986\]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico (Fig. 1). These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria \[@kelloggCharacterizationAerosolizedBacteria2004\], fungi \[@ramirez-camejoFungalDiversitySahara2022\], virus-like particles \[@griffinAfricanDesertDust\], and minerals \[@formenti2003\]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron \[@millsIronPhosphorusColimit2004; @savoieNonseasaltSulfateNitrate1989 ; @graham1982\] . This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups \[@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a].

@@

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours (Fig. 2). *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community [@westrichSaharanDustNutrients2016a]. A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition [@westrichVibrioPopulationDynamics2018a] In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups [@borchardtSaharanDustDeposition2020b]. To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. However, the composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

@@

A high resolution (daily) time series was conducted starting on July 7th and ending on July 19th, 2022 capturing before, during, and after a Saharan dust event. <!--# Add protocol -->

@@

Total *Vibrio* were estimated using genus-specific quantitative polymerase chain reaction (qPCR) [@westrichSaharanDustNutrients2016a]. <!--# Add specific primers, conditions, etc  -->

@@

#### Dust

![](images/paste-0FE55FAA.png)

This figure depicts the summed dust concentration 24hr prior to sample collection (time points 13hr--> 7hr the next day) across the daily time series. During the time series, two small periods of higher dust AOT occurred on the 9th and 13th respectively. A much larger spike in AOT was observed on the 16th.

#### Daily Vibrio

@@

The top portion of the figure depicts total enumerated Vibrio as copies per mL, with color by site. There is a noticeable shift in copies in the Gulf site, a less noticeable but still present shift in Blind Oso Bay, and practically no changes in the Canal site. The bottom portion of the figure depicts the dust concentration (AOT) across the daily time series. Initial shift in copies per mL at the Gulf and Blind Oso occur 24-48hr following initial dust introduction on the 13th, and experience another shift around 24-48hr of higher dust input on the 16th.

#### Temperature & Salinity

![](images/temp_sal.png)

@@

![](images/00000b-01.png){width="406"}

![](images/00000f-01.png){width="404"}

![](images/000009%20(1)-01.png){width="404"}

>>>>

3.  Ultimately, I feel as though I am missing some "oomph"when it comes to my statistical analysis. The linear models seem adequate, but I know that there are multiple variables at play at these sites (and we have the nutrient data for this). I am open to any suggestions that may diversify and expand my analyses!

@@

The Sahara Desert is a significant source of atmospheric dust, eliciting an estimated one billion metric tons each year [@dalmeidaModelSaharanDust1986\]. Annually, plumes of this dust travel across the Atlantic via easterly trade winds to be deposited in the surface waters of the Atlantic, Caribbean, and Gulf of Mexico. These dust events are highly episodic, occurring 3- 4 times a year typically in the summer months and lasting 3-5 days each. Dust aerosols can harbor a wide diversity of bacteria \[@kelloggCharacterizationAerosolizedBacteria2004\], fungi \[@ramirez-camejoFungalDiversitySahara2022\], virus-like particles \[@griffinAfricanDesertDust\], and minerals \[@formenti2003\]. These aerosols also serve as a significant source of macro and micronutrients such as phosphate, nitrate, and iron \[@millsIronPhosphorusColimit2004; @savoieNonseasaltSulfateNitrate1989 ; @graham1982\] . This addition of otherwise limiting resources can lead to rapid and potentially harmful blooms of certain microbial groups \[@lenesSaharanDustPhosphatic2008; @westrichSaharanDustNutrients2016a ; @westrichSaharanDustNutrients2016a].

@@

Previous work has characterized *Vibrio* blooms in response to Saharan dust input in the oligotrophic setting of the Caribbean and subtropical Atlantic (Barbados and Florida Keys, respectively), with surface water concentrations of total *Vibrio* increasing by five to thirty times that found during non-dust conditions and returning to baseline levels within 24-48 hours. *Vibrio* composition within the larger microbial community also shifted following dust deposition, with initial levels of \<1.4% to a peak of 19.8% of the bacterial community [@westrichSaharanDustNutrients2016a]. A similar phenomenon was also observed in the surface waters of the tropical and open ocean mid-Atlantic with *Vibrio* populations increasing 1.5-fold in the mid-Atlantic following deposition [@westrichVibrioPopulationDynamics2018a] In a follow up study in the Florida Keys, episodic dust events during a daily time series promoted a succession of bacterial responses, with declines in *Prochlorococcus* coinciding with initial increases in bacteria belonging to the order *Vibrionales*, followed by subsequent shifts in response of different bacterial groups [@borchardtSaharanDustDeposition2020b]. To date, our understanding of *Vibrio* and other microbial response to dust input has focused on oligotrophic settings, where dust input is considered to be a critical source of limiting nutrients. However, the composition of desert dust can be complex and may have the potential to deliver critical resources that could be exploited by opportunistic microbes (like *Vibrio*) and elicit a growth response despite higher baseline nutrient levels. Given that human exposure is likely in these coastal waters, it is important to elucidate the relationship between dust input and microbial blooms in these higher nutrient environments.

@@

A high resolution (daily) time series was conducted starting on July 7th and ending on July 19th, 2022 capturing before, during, and after a Saharan dust event. Samples were collected in 1L autoclaved Polypropylene Bottles and immediately placed in a cooler on ice. Following transport to the laboratory, the samples were concentrated onto 0.2µm pore size, hydrophobic polycarbonate membranes (25mm in diameter) and stored at -80°C until DNA Extraction using a ZymoBIOMICS DNA Miniprep Kit (Cat#:4300).

@@

Total *Vibrio* concentrations were quantified using a SYBR Green quantitative PCR (qPCR) method. Estimates were made using genus-specific quantitative polymerase chain reaction (qPCR) PCR product using Vibrio group specific primers targeting a variable region of the 16S rRNA gene, 567F, 5′GGCGTAAAGCGCATGCAGGT3′ and 680R, 5′-GAAATTCTACCCCCCTCTACAG-3. Master mix and PCR conditions were derived from [@westrichSaharanDustNutrients2016a]. Briefly, 5µL of 2X SYBR PowerUp Master Mix (Applied Biosystems, Foster City, CA) were added to primers with a final concentration of 0.16µM and PCR water for a final volume of 10µL. Reactions were run in triplicate on a CFX96 Touch Real-Time PCR Detection System (Bio-Rad Laboratories, Hercules, CA) with the following cycling conditions: 2 minutes at 50◦C for UDG activation and 95◦C for 2 min to activate AmpliTaq polymerase and UP, followed by 40 cycles of 95◦C for 3 seconds for denaturation and 60◦C for 30 seconds for annealing and extension. Each run was followed by a dissociation step (60C to 95C by 0.5C increments) to determine a melt curve for analysis of specificity. Each test was run with positive and negative controls in triplicate. Cycle threshold values for each qPCR test was compared to a standard curve representing 10\^1 to 10\^6 gene copies per reaction volume. <!--# Add Std Curve specs -->

@@

#### Dust AOD and Vibrio Enumeration

@@

The top portion of the figure depicts total enumerated Vibrio as copies per mL, with color by site. There is a noticeable shift in copies in the Gulf site, a less noticeable but still present shift in Blind Oso Bay, and practically no changes in the Canal site. The bottom portion of the figure depicts the dust concentration (AOD) across the daily time series. This figure depicts the summed dust concentration 24hr prior to sample collection (time points 13hr--> 7hr the next day) across the daily time series. During the time series, two small periods of higher dust AOD occurred on the 9th and 13th respectively. A much larger spike in AOD was observed on the 16th. Initial shift in copies per mL at the Gulf and Blind Oso occur 24-48hr following initial dust introduction on the 13th, and experience another shift around 24-48hr of higher dust input on the 16th.

#### Environmental Parameters

![](images/000031.png)

@@

![](images/000005-02.png){width="342"}

![](images/000009.png){width="341"}

![](images/00000d.png)
